---
title: 'Applied Econometrics (Part 3)'
subtitle: 'Lviv Data Science Summer School' 
author: "Jozef Barunik & Luboš Hanus"
date: "July 2019"
output:
  html_document:
    fig.retina: 1
    fig_height: 3.5
    fig_width: 5
    keep_md: yes
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    fig_height: 2.1
    fig_width: 4.2
    keep_tex: yes
    latex_engine: xelatex
---
<!-- #### Outline -->
<!-- * Linear AR, MA processes -->
<!-- * VAR analysis -->
<!-- * Connection to Machine Learning techniques -->
<!-- * Macro + Finance application -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#fig.width = 9.5, fig.height = 3.7
```

Let's install required packages first, please change the path

```{r results='hide', message=FALSE, warning=FALSE}
rm(list = ls())
library(data.table)
#library(devtools)
#install_github("wbnicholson/BigVAR/BigVAR")
#install_github("tomaskrehlik/frequencyConnectedness", tag = "0.1.6") 
library(frequencyConnectedness)
library(igraph,warn.conflicts=F)
library(MASS,warn.conflicts=F)
library(BigVAR)
```


# Application: Connectedness
We can further think how to use information in VAR. For example, if we want to measure how the system is connected, FEVDs can be used. 

FEVD gives information on how shocks transmit, how shock in variable $i$ spills over to variable $j$.

This is very useful, as it can be used as a system characterisation

## how to measure connectedness of a system?

We will work with 11 most important banks data, returns and volatility (measure of risk, uncertainty)

```{r fig.width = 9.5, fig.height = 9.5}

load(file="Example_banks.RData")
## Example of data
plot.ts(ret[,2])
lines(sqrt(rv[,2]),col="red")

```

Consider an $N$-dimensional vector $\mathbf{RV_t} = (RV_{1t},\ldots,RV_{nt})'$ holding $N$ assets that we model by VAR($p$) system as

$$\mathbf{RV_t} = \sum_{i=1}^p \mathbf{\Phi}_i \mathbf{RV}_{t-i}+ \boldsymbol{\epsilon}_t$$

with $\boldsymbol{\epsilon}_t\sim N(0,\mathbf{\Sigma}_{\epsilon})$. Moving average representation: 

$$\mathbf{RV}_t = \sum_{i=0}^{\infty}\mathbf{\Psi}_{i}\boldsymbol{\epsilon}_{t-i}.$$

These coefficients are extremely useful, since they show how shocks from variables transmit in the system, hence how the system is connected. The onlyt problem is identification of shocks.

We can use Generalized FEVDs
$$\omega_{ij}^H=\frac{\sigma_{jj}^{-1}\sum_{h=0}^{H-1}\left( \mathbf{e}'_i \mathbf{\Psi}_h \mathbf{\Sigma}_{\epsilon}\mathbf{e}_j \right)^2}{\sum_{h=0}^{H-1}\left( \mathbf{e}'_i \mathbf{\Psi}_h \mathbf{\Sigma}_{\epsilon}\mathbf{\Psi}'_h\mathbf{e}_i \right)},$$

where $\mathbf{e}$ is a selection vector with one for $i$th element, zero otherwise. The sum of elements in rows is not equal to one, so we normalize it

$$ \widetilde{\omega}_{ij}^H = \frac{\omega_{ij}^H}{\sum_{j=1}^N \omega_{ij}^H}.$$

and finally obtain measure of connectedness $\mathcal{C} \in (0,100)$ 
$$\mathcal{C}^H=100\times \frac{1}{N} \sum_{\substack{i,j=1\\ i\ne j}}^N\widetilde{\omega}_{ij}^H.$$

Let's have a look at the connectedness table (elements of teh table are $\widetilde{\omega}_{ij}^{10}$)

```{r}
RV <- log(sqrt(rv))
con<-spilloverDY12(VAR(RV, p = 4, type = "const"), n.ahead = 10, no.corr=F)
con
```
Note FROM, and TO which are simple sums, and tell us *directional connectedness*

Total connectedness is
```{r}
overall(con)
```

## Dynamic connectedness

We can easily compute rolling window dynamics. 

NOTE: this takes time!
```{r fig.width = 9.5, fig.height = 9.5, cache=TRUE}
spills<-spilloverRollingDY12(data = RV, n.ahead = 10, no.corr = F, "VAR", params_est = list(p = 2, type = "const"), window = 250)
plot.ts(as.numeric(overall(spills)[[1]]), lwd=0.5,ylim=c(50,100),xlab="",ylab="")

```

## Network representation

We can also note that VAR can be represented as a network

```{r fig.width = 9.5, fig.height = 9.5, cache=TRUE}
tab<-spilloverDY12(VAR(RV[1600:2000], p = 4, type = "const"), n.ahead = 10, no.corr=F)

adj=t(100*tab$tables[[1]])
diag(adj)=0  # 0 in an adjacency matrix means no edge is to be created
adj[adj<=9]=0

netw=graph_from_adjacency_matrix(adj,mode="directed",weighted=T)
plot(netw,
     layout=layout_with_fr,
     vertex.size=to(tab)[[1]]*5
)
```


## Machine learning connectedness
In case we like to compute connectedness of large system, we can use lasso again.

```{r fig.width = 9.5, fig.height = 9.5, message=FALSE, warning=FALSE, cache=TRUE}
# Set very small lambda (gran parameter) to mimic no penalty
Model1 = constructModel(as.matrix(RV[1600:2000]), p = 4, struct = "Basic", gran=c(2), verbose = F, VARX = list(), ownlambdas = T)
Model1Results = cv.BigVAR(Model1)

# spillover table from BigVaR
tabBig<-spilloverDY12(Model1Results, n.ahead = 10, no.corr = F)

adj=t(100*tabBig$tables[[1]])
diag(adj)=0  # 0 in an adjacency matrix means no edge is to be created
adj[adj<=9]=0

netw=graph_from_adjacency_matrix(adj,mode="directed",weighted=T)
plot(netw,
     layout=layout_with_fr,
     vertex.size=to(tabBig)[[1]]*5
)
```

